---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Dashiell Ridolfi-Starr"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse); library(here); library(agricolae)
getwd(); here()

lake.chemphys.data <- read.csv(
  here("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)
lake.chemphys.data$sampledate <- mdy(lake.chemphys.data$sampledate)

#2
mytheme <-
  theme_gray() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no relationship between mean July temperature and depth in all lakes.
Ha: There is a relationship between mean July temperature and depth in all lakes.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
chemphys.wrangled <-
  lake.chemphys.data %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum,depth,temperature_C) %>%
  na.omit()

#5
ggplot(chemphys.wrangled, aes(x = depth, y = temperature_C)) +
  geom_point(color = "#76a5af", alpha = 0.5) +
  geom_smooth(method = "lm",color = "black") +
  ylim(0,35) +
  labs(
    x = "Depth (m)",
    y = "Temperature (°C)",
    title = "July Lake Temperatures by Depth")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This plot suggests that overall, temperature decreases with depth. The data's distribution suggests a non-linear relationship. The rate of change in temperature values appears slow at first, then rapid between roughly 2.5-7.5 m of depth, then slow again, with very little change after approximately 10 m. Variability in temperature also appears to decrease with depth. 

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
temp.depth.regression <- lm(
  data = chemphys.wrangled,
  temperature_C ~ depth)
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: There is almost certainly a strong relationship between temperature and depth. Depth explains approximately 73.9% of the variance in temperature. This is based on 9726 degrees of freedom and has a highly statistically significant p-value of 2.2e-16. Transforming the data may result in a more linear distribution and better fit.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp.multiple.regression <- lm(
  data = chemphys.wrangled,
  temperature_C ~ year4 + daynum + depth
  )
step(temp.multiple.regression)

#10
summary(temp.multiple.regression) #no change in explanatory variables

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables consists of year, day number, and depth, which explain approximately 74.1% of variance in temperature. This is only a very slight improvement over the single regression model. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#ANOVA model
temps.anova1 <- aov(
  data = chemphys.wrangled,
  temperature_C ~ lakename
  )
summary(temps.anova1)

#linear model
temps.anova2 <- lm(
  data = chemphys.wrangled,
  temperature_C ~ lakename
  )
summary(temps.anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a highly statistically significant difference in mean temperature among the lakes, although the R-squared value is very low; the lake name explains only about 3.9% of the variance in mean temperature.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2a, fig.height=4}
#14.
ggplot(chemphys.wrangled, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha= 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0,35) +
  labs(
    x = "Depth (m)",
    y = "Temperature (°C)",
    title = "July Lake Temperatures ",
    subtitle = "by Depth and Lake Name",
    color = "Lake Name") +
  theme(plot.subtitle = element_text(hjust = 0.5))
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temps.anova1)

temp.groups <- HSD.test(temps.anova1, "lakename", group = TRUE)
temp.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to Tukey's HSD test, the lakes with the same mean temperature as Peter Lake are Ward Lake and Paul Lake. There is no lake with a statistically distinct mean temperature. 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Since we are investigating the effects of a binary categorical variable a continuous response, we could use a two-sample t-test to compare mean temperatures.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
cramptonward.subset <-
  chemphys.wrangled %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

cramptonward.ttest <- t.test(
  cramptonward.subset$temperature_C ~ cramptonward.subset$lakename
)
cramptonward.ttest

```

>Answer: The t-test produced a non-statistically significant p-value of 0.2649. This means that the mean July temperatures of Ward Lake and Crampton Lake are not statistically different. This agrees with the results of Tukey's HSD test. 
